---
title: "Maryland Socioeconomic Trends Dashboard - Population, Households, and Employment (MSTM)"
runtime: shiny
author: ""
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    self_contained: true
    social: menu
    theme: sandstone
---

```{r setup, include=FALSE}
# Load required packages
library(flexdashboard)
library(sf)
library(dplyr)
library(tidyr)
library(leaflet)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(readxl)
library(rlang)    # for dynamic column names

# Turn off clutter
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(fig.height = 5, fig.width = 10)

# --- Utility function for Indian-style comma formatting ---
comma_format_indian <- function(x) {
  format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
}


# Clean, bold look for all charts
theme_set(theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title = element_text(size = 12),
    panel.grid.minor = element_blank()
  ))
theme_update(
  axis.title.x = element_text(size = 13, face = "bold", margin = margin(t = 10)),
  axis.title.y = element_text(size = 13, face = "bold", margin = margin(r = 10)),
  axis.text = element_text(size = 11)
)
options(plotly.layout.margin = list(l = 120, r = 60, t = 80, b = 80))

# --- File paths ---
base_path  <- "SE_Data"
zones_path <- file.path(base_path, "Zones.shp")
excel_path <- "Output_L2.xlsx"

# --- Load preprocessed shapefile ---
zones_poly <- readRDS("zones_ready.rds")

# --- Compute area in square kilometers ---
zones_poly <- zones_poly %>%
  mutate(AREA_KM2 = as.numeric(st_area(geometry)) / 1e6)

# --- Compute area in square kilometers for each zone ---
zones_poly <- zones_poly %>%
  mutate(AREA_KM2 = as.numeric(st_area(geometry)) / 1e6)



# Simplify geometry to speed up rendering (fails gracefully on error)
zones_poly <- tryCatch(
  st_simplify(zones_poly, dTolerance = 200),
  error = function(e) zones_poly
)

# Restrict to Maryland zones using available attributes
if ("State" %in% names(zones_poly)) {
  zones_poly <- zones_poly %>% filter(State == "Maryland")
} else {
  zones_poly <- zones_poly %>%
    filter(grepl("MD", County) | grepl("Maryland", REGION, ignore.case = TRUE))
}

# Keep only valid polygons
zones_poly <- zones_poly %>%
  filter(st_geometry_type(.) %in% c("POLYGON", "MULTIPOLYGON")) %>%
  st_make_valid()

# Compute land area in square kilometers
zones_poly$Area_km2 <- as.numeric(st_area(zones_poly)) / 1e6

# Convert numeric fields in shapefile (base data)
base_numeric <- c("POP", "HH", "Emp_Tot", "Emp_Ret", "Emp_Ind", "Emp_Off",
                  "Emp_Edu", "Emp_Hea", "Emp_Foo", "Emp_Oth")
zones_poly <- zones_poly %>%
  mutate(across(all_of(base_numeric), ~ suppressWarnings(as.numeric(gsub(",", "", trimws(as.character(.x)))))))

# Rename region codes to descriptive names
zones_poly$REGION <- recode(zones_poly$REGION,
  "SMZ_WEST"         = "Western Maryland",
  "SMZ_MWCOG"        = "DC Metro Area",
  "SMZ_EASTERN/PENN" = "Eastern Shore / PA Border",
  "SMZ_EAST"         = "Central East",
  "SMZ_DC"           = "Washington Corridor",
  "SMZ_BMC"          = "Baltimore Metro",
  "RMZ"              = "Rural Maryland",
  "HALO_WV"          = "Halo WV",
  "HALO_VA"          = "Halo VA",
  "HALO_PENN"        = "Halo PA",
  "HALO_PA"          = "Halo North PA",
  "HALO_MWCOG VA"    = "Halo DC-VA"
)

# --- Load socioeconomic data from Excel ---
mstm_data <- read_excel(excel_path)
# Standardize column names for easier matching
names(mstm_data) <- gsub("\\.", "_", names(mstm_data))

```

```{r}
# 🧭 Year selector UI
inputPanel(
  selectInput(
    "year_choice",
    "Select Year:",
    choices = c(2020, 2023, 2025, 2030, 2035, 2040, 2045, 2050),
    selected = 2023
  )
)
```

```{r}
zones_reactive <- reactive({
  req(input$year_choice)
  isolate({
    # Convert actual year (e.g., 2023) into Excel suffix (e.g., "23")
    year_suffix <- case_when(
      input$year_choice == 2020 ~ "20",
      input$year_choice == 2023 ~ "23",
      input$year_choice == 2025 ~ "25",
      input$year_choice == 2030 ~ "30",
      input$year_choice == 2035 ~ "35",
      input$year_choice == 2040 ~ "40",
      input$year_choice == 2045 ~ "45",
      input$year_choice == 2050 ~ "50",
      TRUE ~ "23"
    )

    df <- zones_poly %>%
      left_join(mstm_data, by = c("LVL_2_ID" = "L2_TAZ_MST")) %>%
      mutate(
        POP     = !!sym(paste0("HPOP_", year_suffix)),
        HH      = !!sym(paste0("HH_",   year_suffix)),
        Emp_Tot = !!sym(paste0("EMP_",  year_suffix)),
        Emp_Ind = !!sym(paste0("IND_",  year_suffix)),
        Emp_Ret = !!sym(paste0("RET_",  year_suffix)),
        Emp_Off = !!sym(paste0("SER_",  year_suffix)),
        Emp_Edu = !!sym(paste0("OTHED_", year_suffix)),
        Emp_Hea = !!sym(paste0("OTHHSP_", year_suffix)),
        Emp_Foo = !!sym(paste0("OTHFD_", year_suffix)),
        Emp_Oth = !!sym(paste0("OTHOTH_", year_suffix))
      ) %>%
      mutate(
        # Derived metrics
        Emp_to_Pop_Ratio = ifelse(POP > 0, (Emp_Tot / POP) * 100, NA_real_),
        Avg_HH_Size      = ifelse(HH > 0, POP / HH, NA_real_),
        POP_Density      = ifelse(Area_km2 > 0, POP / Area_km2, NA_real_),
        HH_Density       = ifelse(Area_km2 > 0, HH  / Area_km2, NA_real_),
        EMP_Density      = ifelse(Area_km2 > 0, Emp_Tot / Area_km2, NA_real_)
      )

    return(df)
  })
})

```

# Population Patterns {.tabset}
## Charts {.tabset}
### Region Map 
```{r, fig.height=5}
renderLeaflet({
  input$year_choice
  zones_df <- zones_reactive()
  zones_df$POP_SQRT <- sqrt(zones_df$POP)

  pal_region <- colorNumeric(
    palette = colorRampPalette(c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"))(8),
    domain  = zones_df$POP_SQRT,
    na.color = "transparent"
  )

  leaflet(zones_df, options = leafletOptions(minZoom = 6, zoomControl = TRUE)) %>%
    addProviderTiles(providers$Esri.WorldTopoMap, group = "Topographic (City Labels)") %>%
    addProviderTiles(providers$Esri.WorldStreetMap, group = "Street View") %>%
    addProviderTiles(providers$CartoDB.Positron, group = "Light Mode") %>%
    addPolygons(
      data = zones_df %>% filter(!is.na(POP) & POP > 0),
      fillColor   = ~pal_region(sqrt(POP)),
      weight      = 0.4,
      color       = "#444444",
      fillOpacity = 0.6,
      label       = ~sprintf(
        "<b>County:</b> %s<br>\n<b>Region:</b> %s<br>\n<b>Total Population:</b> %s",
        County, REGION, format(POP, big.mark = ",")
      ) %>% lapply(htmltools::HTML),
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#000000",
        fillOpacity = 0.8,
        bringToFront = TRUE
      )
    ) %>%
    addLegend(
      position = "bottomright",
      pal      = pal_region,
      values   = ~sqrt(POP),
      title    = htmltools::HTML("<b>Population Intensity</b><br><span style='font-size:11px;'>Lighter = smaller population<br>Darker = denser zones</span>"),
      opacity  = 0.85,
      labFormat = labelFormat(big.mark = ",")
    ) %>%
    addLayersControl(
      baseGroups = c("Topographic (City Labels)", "Street View", "Light Mode"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    setView(lng = -76.7, lat = 39.0, zoom = 8)
})

```

### County-Level Population Map {.tabset}
```{r}
renderLeaflet({
  # Dissolve zones into counties and compute total population per county
  input$year_choice
  zones_df <- zones_reactive()

  pop_by_county <- zones_df %>%
    st_make_valid() %>%
    group_by(County) %>%
    summarise(Total_POP = sum(POP, na.rm = TRUE), .groups = "drop") %>%
    st_union(by_feature = TRUE) %>%
    st_buffer(0) %>%
    st_as_sf()

  # Palette for counties
  pop_by_county$POP_SQRT <- sqrt(pop_by_county$Total_POP)
  pal_county <- colorNumeric(
    colorRampPalette(c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"))(8),
    domain = pop_by_county$POP_SQRT,
    na.color = "transparent"
  )

  leaflet(pop_by_county, options = leafletOptions(minZoom = 6, zoomControl = TRUE)) %>%
    addProviderTiles(providers$Esri.WorldTopoMap,   group = "Topographic (City Labels)") %>%
    addProviderTiles(providers$Esri.WorldStreetMap, group = "Street View") %>%
    addProviderTiles(providers$CartoDB.Positron,    group = "Light Mode") %>%
    addPolygons(
      fillColor   = ~pal_county(POP_SQRT),
      fillOpacity = 0.72,
      color       = "#2b2b2b",
      weight      = 1.1,
      label       = ~sprintf("<b>%s County</b><br><b>Total Population:</b> %s",
                             County, format(Total_POP, big.mark = ",")) %>% lapply(htmltools::HTML),
      highlightOptions = highlightOptions(weight = 2, color = "#000000", fillOpacity = 0.9, bringToFront = TRUE)
    ) %>%
    addLegend(
      "bottomright", pal = pal_county, values = ~POP_SQRT, opacity = 0.85,
      title = htmltools::HTML("<b>Total Population</b><br><span style='font-size:11px;'>Lighter = lower<br>Darker = higher</span>"),
      labFormat = labelFormat(big.mark = ",")
    ) %>%
    addLayersControl(
      baseGroups = c("Topographic (City Labels)", "Street View", "Light Mode"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    setView(lng = -76.7, lat = 39.0, zoom = 8)
})

```

## Row
### Population and Employment Density by Region
```{r}
renderPlotly({
  input$year_choice

  density_region <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(
      Pop_Density = sum(POP, na.rm = TRUE) / sum(AREA_KM2, na.rm = TRUE),
      Emp_Density = sum(Emp_Tot, na.rm = TRUE) / sum(AREA_KM2, na.rm = TRUE)
    ) %>%
    pivot_longer(cols = c(Pop_Density, Emp_Density),
                 names_to = "Type", values_to = "Density") %>%
    mutate(Type = recode(Type,
                         "Pop_Density" = "Population Density",
                         "Emp_Density" = "Employment Density"))

  p_density <- ggplot(density_region, aes(
    x = reorder(REGION, Density),
    y = Density,
    fill = Type,
    text = paste0(
      "<b>Region:</b> ", REGION,
      "<br><b>Type:</b> ", Type,
      "<br><b>Density:</b> ", round(Density, 1), " per km²"
    )
  )) +
    geom_col(position = "dodge", alpha = 0.9) +
    coord_flip() +
    scale_y_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +
    labs(
      title = paste0("Population & Employment Density by Region – ", input$year_choice),
      subtitle = "Densities normalized by area (per km²)",
      x = "Region",
      y = "Density (per km²)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )

  ggplotly(p_density, tooltip = "text") %>%
    layout(
      title = list(
        text = paste0("<b>Population & Employment Density by Region – ", input$year_choice, "</b>"),
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Density (per km²)", font = list(size = 13)), tickformat = ",f"),
      yaxis = list(title = list(text = "Region", font = list(size = 13)), autorange = "reversed"),
      margin = list(l = 140, r = 50, t = 100, b = 100)
    )
})
```


## Row
### Total Population & Household by Region
```{r}
renderPlotly({
  input$year_choice
  
  pop_sum <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(
      Total_POP = sum(POP, na.rm = TRUE),
      Total_HH  = sum(HH,  na.rm = TRUE)
    ) %>%
    pivot_longer(
      cols = c("Total_POP", "Total_HH"),
      names_to = "Type",
      values_to = "Value"
    )
  
  # --- Force full comma formatting on ggplot axis ---
  p_total <- ggplot(pop_sum, aes(x = REGION, y = Value, fill = Type)) +
    geom_col(position = "dodge", alpha = 0.9) +
    scale_y_continuous(
      labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)
    ) +   # ✅ Forces full comma format (no 1e+06)
    labs(
      title = "Total Population and Households by Region",
      subtitle = "Aggregated totals better reflect Maryland’s regional distribution",
      x = "Region",
      y = "Total Value"
    ) +
    scale_fill_brewer(palette = "Set2") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  
  # --- Ensure commas in Plotly display ---
  ggplotly(p_total, tooltip = "text", height = 520) %>%
    layout(
      title = list(
        text = "<b>Total Population and Household by Region</b><br><span style='font-size:12px;color:gray'>Total household and population values per Maryland region</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Region", font = list(size = 13))),
      yaxis = list(
        title = list(text = "Population", font = list(size = 13)),
        tickformat = ",f"   # ✅ Forces full comma style in Plotly axis too
      ),
      margin = list(l = 160, r = 40, t = 90, b = 240)
    )
})

```

### Population by County
```{r}
renderPlotly({
  input$year_choice
  
  pop_county <- st_drop_geometry(zones_reactive()) %>%
    group_by(County) %>%
    summarise(Total_POP = sum(POP, na.rm = TRUE), .groups = "drop") %>%
    filter(Total_POP > 0) %>%
    arrange(desc(Total_POP))

  p_pop_county <- ggplot(
    pop_county %>% arrange(desc(Total_POP)),
    aes(
      x = reorder(County, Total_POP),
      y = Total_POP,
      fill = County,
      text = paste(
        "<b>County:</b>", County,
        "<br><b>Total Population:</b>", format(Total_POP, big.mark = ",", scientific = FALSE)
      )
    )
  ) +
    geom_col(show.legend = FALSE, alpha = 0.9) +
    coord_flip() +
    scale_x_discrete(limits = rev(unique(pop_county$County))) +
    scale_y_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +  # ✅ format axis with commas
    labs(
      title = "Total Population by County",
      subtitle = "Counties sorted in descending order (highest at top)",
      x = "County",
      y = "Total Population"
    ) +
    theme_minimal(base_size = 12)

  # Render interactive plot
  ggplotly(p_pop_county, tooltip = "text", height = 700) %>%
    layout(
      title = list(
        text = "<b>Total Population by County</b><br><span style='font-size:12px;color:gray'>Highest population counties appear at the top</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(
        title = list(text = "Total Population", font = list(size = 13)),
        tickformat = ",f"   # ✅ comma format for population values
      ),
      yaxis = list(title = list(text = "County", font = list(size = 13))),
      margin = list(l = 160, r = 40, t = 90, b = 340)
    )
})


```

## Row{.full}
### Total Population by Region
```{r}
renderPlotly({
  input$year_choice
  
  # Compute total population by region
  df_pop_tot <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(Total_POP = sum(POP, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(Total_POP))

  p_pop_tot <- ggplot(
    df_pop_tot,
    aes(
      x = reorder(REGION, Total_POP),
      y = Total_POP,
      fill = REGION,
      text = paste0(
        "<b>Region:</b> ", REGION,
        "<br><b>Total Population:</b> ", format(Total_POP, big.mark = ",", scientific = FALSE)
      )
    )
  ) +
    geom_col(alpha = 0.9, width = 0.55, show.legend = TRUE) +
    coord_flip() +
    # ✅ Force comma formatting for all y-axis values
    scale_y_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +
    labs(
      title = "Total Population by Region",
      subtitle = "",
      x = "Region",
      y = "Total Population"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.margin = margin(20, 40, 40, 100),
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      axis.title.x = element_text(size = 12, face = "bold", margin = margin(t = 15)),
      axis.title.y = element_text(size = 12, face = "bold", margin = margin(r = 15))
    )

  # ✅ Render interactive plot with commas in Plotly axis
  ggplotly(p_pop_tot, tooltip = "text", height = 550) %>%
    layout(
      title = list(
        text = paste0(
          "<b>Total Population by Region</b><br>",
          "<span style='font-size:12px;color:gray'></span>"
        ),
        xanchor = "center",
        x = 0.5
      ),
      xaxis = list(
        title = list(text = "Total Population", font = list(size = 13)),
        tickformat = ",f",   # ✅ Comma formatting
        side = "bottom",
        automargin = TRUE
      ),
      yaxis = list(
        title = list(text = "Region", font = list(size = 13)),
        automargin = TRUE,
        anchor = "x"
      ),
      margin = list(l = 160, r = 60, t = 100, b = 150)
    )
})

```

## Row
### Most Populated County in Each Region
```{r}
renderPlotly({
  input$year_choice
  
  # Identify the most populated county in each region
  top_counties_region <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION, County) %>%
    summarise(Total_POP = sum(POP, na.rm = TRUE), .groups = "drop") %>%
    group_by(REGION) %>%
    slice_max(order_by = Total_POP, n = 1, with_ties = FALSE) %>%
    ungroup()

  p_top_counties_region <- ggplot(
    top_counties_region,
    aes(
      x = reorder(REGION, Total_POP),
      y = Total_POP,
      fill = REGION,
      text = paste0(
        "<b>Region:</b> ", REGION,
        "<br><b>Top County:</b> ", County,
        "<br><b>Total Population:</b> ", format(Total_POP, big.mark = ",", scientific = FALSE)
      )
    )
  ) +
    geom_col(alpha = 0.9, width = 0.6) +
    coord_flip() +
    # ✅ Add comma formatting for numeric axis
    scale_y_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +
    labs(
      title = paste0("Most Populated County in Each Maryland Region – ", input$year_choice),
      subtitle = "Top county in each region based on total population",
      x = "Region",
      y = "Total Population"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      axis.text.y = element_text(face = "bold", size = 12)
    )

  # ✅ Interactive plot with comma tick formatting
  ggplotly(p_top_counties_region, tooltip = "text") %>%
    layout(
      title = list(
        text = paste0(
          "<b>Most Populated County in Each Maryland Region – ",
          input$year_choice,
          "</b><br><span style='font-size:12px;color:gray'>Top county in each region based on population</span>"
        ),
        xanchor = "center", x = 0.5
      ),
      xaxis = list(
        title = list(text = "Total Population", font = list(size = 13)),
        tickformat = ",f"   # ✅ Full comma-based number formatting
      ),
      yaxis = list(title = list(text = "Region", font = list(size = 13))),
      margin = list(l = 120, r = 40, t = 100, b = 100)
    )
})
```


# Household Dynamics {.tabset}

## Charts {.tabset}
### Household Map
```{r, fig.height=5}
renderLeaflet({
  input$year_choice
  # Prepare data for household map
  zones_df <- zones_reactive()
  zones_df$HH_SQRT <- sqrt(zones_df$HH)

  pal_hh <- colorNumeric(
    palette = colorRampPalette(c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"))(8),
    domain  = zones_df$HH_SQRT,
    na.color = "transparent"
  )

  leaflet(zones_df, options = leafletOptions(minZoom = 6, zoomControl = TRUE)) %>%
    addProviderTiles(providers$Esri.WorldTopoMap, group = "Topographic (City Labels)") %>%
    addProviderTiles(providers$Esri.WorldStreetMap, group = "Street View") %>%
    addProviderTiles(providers$CartoDB.Positron, group = "Light Mode") %>%
    addPolygons(
      data = zones_df %>% filter(!is.na(HH) & HH > 0),
      fillColor   = ~pal_hh(sqrt(HH)),
      weight      = 0.4,
      color       = "#444444",
      fillOpacity = 0.6,
      label       = ~sprintf(
        "<b>Zone:</b> %s<br>\n<b>County:</b> %s<br>\n<b>Households:</b> %s",
        GIS_Zone, County, format(HH, big.mark = ",")
      ) %>% lapply(htmltools::HTML),
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#000000",
        fillOpacity = 0.8,
        bringToFront = TRUE
      )
    ) %>%
    addLegend(
      position = "bottomright",
      pal      = pal_hh,
      values   = ~sqrt(HH),
      title    = htmltools::HTML("<b>Total Households</b><br><span style='font-size:11px;'>Lighter = fewer households<br>Darker = denser zones</span>"),
      opacity  = 0.85,
      labFormat = labelFormat(big.mark = ",")
    ) %>%
    addLayersControl(
      baseGroups = c("Topographic (City Labels)", "Street View", "Light Mode"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    setView(lng = -76.7, lat = 39.0, zoom = 8)
})

```

### County-Level Household Map
```{r}
### County-Level Household Map
renderLeaflet({
  input$year_choice
  # Aggregate households by county
  df <- zones_reactive()

  hh_by_county <- df %>%
    st_make_valid() %>%
    group_by(County) %>%
    summarise(Total_HH = sum(HH, na.rm = TRUE), .groups = "drop") %>%
    st_union(by_feature = TRUE) %>%
    st_buffer(0) %>%
    st_as_sf()

  # Palette for household counts
  hh_by_county$HH_SQRT <- sqrt(hh_by_county$Total_HH)
  pal_hh_county <- colorNumeric(
    colorRampPalette(c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"))(8),
    domain = hh_by_county$HH_SQRT,
    na.color = "transparent"
  )

  leaflet(hh_by_county, options = leafletOptions(minZoom = 6, zoomControl = TRUE)) %>%
    addProviderTiles(providers$Esri.WorldTopoMap,   group = "Topographic (City Labels)") %>%
    addProviderTiles(providers$Esri.WorldStreetMap, group = "Street View") %>%
    addProviderTiles(providers$CartoDB.Positron,    group = "Light Mode") %>%
    addPolygons(
      fillColor   = ~pal_hh_county(HH_SQRT),
      fillOpacity = 0.72,
      color       = "#2b2b2b",
      weight      = 1.1,
      label       = ~sprintf("<b>%s County</b><br><b>Total Households:</b> %s",
                             County, format(Total_HH, big.mark = ",")) %>% lapply(htmltools::HTML),
      highlightOptions = highlightOptions(weight = 2, color = "#000000", fillOpacity = 0.9, bringToFront = TRUE)
    ) %>%
    addLegend(
      "bottomright", pal = pal_hh_county, values = ~HH_SQRT, opacity = 0.85,
      title = htmltools::HTML("<b>Total Households</b><br><span style='font-size:11px;'>Lighter = lower<br>Darker = higher</span>"),
      labFormat = labelFormat(big.mark = ",")
    ) %>%
    addLayersControl(
      baseGroups = c("Topographic (City Labels)", "Street View", "Light Mode"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    setView(lng = -76.7, lat = 39.0, zoom = 8)
})

```

## Row
### Average Household Density by Region
```{r}
renderPlotly({
  input$year_choice
  # Compute average household density per region
  avg_hh_density <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(Mean_HH_Density = mean(HH_Density, na.rm = TRUE)) %>%
    arrange(desc(Mean_HH_Density))

  p_hh_density <- ggplot(avg_hh_density, aes(
    x = reorder(REGION, Mean_HH_Density),
    y = Mean_HH_Density,
    fill = REGION,
    text = paste("Region:", REGION, "<br>Avg. Household Density:", round(Mean_HH_Density, 1), "per km²")
  )) +
    geom_col(width = 0.7) +
    coord_flip() +
    labs(
      title = "Average Household Density by Region",
      subtitle = "Each value shows average number of households per square kilometer",
      x = "Region",
      y = "Avg. Households per km²"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30")
    ) +
    guides(fill = "none")

  # Render interactive plot
  ggplotly(p_hh_density, tooltip = "text") %>%
    layout(
      title = list(
        text = "<b>Average Household Density by Region</b><br><span style='font-size:12px;color:gray'>Avg. households per km²</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Avg. Households per km²", font = list(size = 13))),
      yaxis = list(title = list(text = "Region", font = list(size = 13)), autorange = "reversed"),
      margin = list(l = 120, r = 40, t = 90, b = 80)
    )
})

```

## Row
### Average Households by Region
```{r}
renderPlotly({
  input$year_choice
  # Compute average number of households per zone in each region
  avg_hh <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(Mean_HH = mean(HH, na.rm = TRUE)) %>%
    arrange(desc(Mean_HH))

  p_avg_hh <- ggplot(avg_hh, aes(
    x = reorder(REGION, Mean_HH),
    y = Mean_HH,
    fill = REGION,
    text = paste(
      "<b>Region:</b>", REGION,
      "<br><b>Avg. Households per Zone:</b>", scales::comma(round(Mean_HH, 0))
    )
  )) +
    geom_col(alpha = 0.9) +
    coord_flip() +
    labs(
      title = "Average Household Size by Region",
      subtitle = "Represents the average number of households per planning zone",
      x = "Region",
      y = "Avg. Households per Zone"
    ) +
    guides(fill = "none") +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
      axis.title.y = element_text(size = 13),
      axis.title.x = element_text(size = 13)
    )

  # Render interactive plot
  ggplotly(p_avg_hh, tooltip = "text") %>%
    layout(
      title = list(
        text = "<b>Average Household Size by Region</b><br><span style='font-size:12px;color:gray'>Each value represents the mean number of households per zone within a region</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Avg. Households per Zone", font = list(size = 13))),
      yaxis = list(title = list(text = "Region", font = list(size = 13)), autorange = "reversed"),
      margin = list(l = 120, r = 40, t = 90, b = 80)
    )
})

```

## Row{.full}
### Frequency of Household Size Groups
```{r}
renderPlotly({
  input$year_choice
  # Categorise zones by average household size into groups
  zone_df <- st_drop_geometry(zones_reactive()) %>%
    mutate(HHSIZE_group = case_when(
      Avg_HH_Size <= 1.5 ~ "1",
      Avg_HH_Size <= 2.5 ~ "2",
      Avg_HH_Size <= 3.5 ~ "3",
      Avg_HH_Size <= 4.5 ~ "4",
      Avg_HH_Size <= 5.5 ~ "5",
      TRUE               ~ "6"
    ))

  # Define colours and labels for each group
  hh_colors <- c(
    "1" = "#1f77b4",  # Single / Very Small
    "2" = "#ff7f0e",  # Small
    "3" = "#2ca02c",  # Medium
    "4" = "#d62728",  # Large
    "5" = "#9467bd",  # Very Large
    "6" = "#8c564b"   # Extended
  )

  hh_labels <- c(
    "1" = "Single / Very Small (0–1.5 people)",
    "2" = "Small (1.5–2.5 people)",
    "3" = "Medium (2.5–3.5 people)",
    "4" = "Large (3.5–4.5 people)",
    "5" = "Very Large (4.5–5.5 people)",
    "6" = "Extended (5.5+ people)"
  )

  # Plot grouped frequencies
  p_hh_group <- ggplot(zone_df, aes(x = HHSIZE_group, fill = HHSIZE_group)) +
    geom_bar(color = "black", alpha = 0.9) +
    scale_fill_manual(values = hh_colors, name = "Household Size Group", labels = hh_labels) +
    labs(
      title = "Frequency of Household Size Groups",
      subtitle = "Number of Maryland planning zones within each average household size range",
      x = "Average Household Size Group",
      y = "Number of Zones"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
      axis.text.x = element_text(size = 11),
      legend.position = "right"
    )

  # Render interactive plot
  ggplotly(p_hh_group, tooltip = c("x", "y")) %>%
    layout(
      title = list(
        text = "<b>Frequency of Household Size Groups</b><br><span style='font-size:12px;color:gray'></span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Average Household Size Group", font = list(size = 13))),
      yaxis = list(title = list(text = "Number of Zones", font = list(size = 13))),
      margin = list(l = 90, r = 40, t = 100, b = 90)
    )
})

```

### Household Count Distribution
```{r}
renderPlotly({
  input$year_choice
  # Plot distribution of household counts per zone
  p_hh_density <- ggplot(st_drop_geometry(zones_reactive()), aes(x = HH)) +
    geom_density(fill = "steelblue", alpha = 0.4) +
    geom_histogram(aes(y = after_stat(density)), binwidth = 100, fill = "orange", color = "black", alpha = 0.5) +
    labs(
      title = "Household Count Distribution",
      subtitle = "Shows the spread and concentration of household counts across all Maryland zones",
      x = "Households per Zone",
      y = "Density"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30")
    )

  # Render interactive plot
  ggplotly(p_hh_density, tooltip = "x") %>%
    layout(
      title = list(
        text = "<b>Household Count Distribution</b><br><span style='font-size:12px;color:gray'>Distribution of household counts across all zones</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Number of Households", font = list(size = 13)), tickformat = ","),
      yaxis = list(title = list(text = "Density", font = list(size = 13))),
      margin = list(l = 100, r = 50, t = 90, b = 100)
    )
})

```

# Employment Structure {.tabset}

## Charts {.tabset}
### Employment Map
```{r, fig.height=5}
renderLeaflet({
  input$year_choice
  # Prepare data for employment map
  zones_df <- zones_reactive()
  zones_df$EMP_SQRT <- sqrt(zones_df$Emp_Tot)

  pal_emp <- colorNumeric(
    palette = colorRampPalette(c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"))(8),
    domain  = zones_df$EMP_SQRT,
    na.color = "transparent"
  )

  leaflet(zones_df, options = leafletOptions(minZoom = 6, zoomControl = TRUE)) %>%
    addProviderTiles(providers$Esri.WorldTopoMap, group = "Topographic (City Labels)") %>%
    addProviderTiles(providers$Esri.WorldStreetMap, group = "Street View") %>%
    addProviderTiles(providers$CartoDB.Positron, group = "Light Mode") %>%
    addPolygons(
      data = zones_df %>% filter(!is.na(Emp_Tot) & Emp_Tot > 0),
      fillColor   = ~pal_emp(sqrt(Emp_Tot)),
      weight      = 0.4,
      color       = "#444444",
      fillOpacity = 0.6,
      label       = ~sprintf(
        "<b>Zone:</b> %s<br>\n<b>County:</b> %s<br>\n<b>Total Employment:</b> %s",
        GIS_Zone, County, format(Emp_Tot, big.mark = ",")
      ) %>% lapply(htmltools::HTML),
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#000000",
        fillOpacity = 0.8,
        bringToFront = TRUE
      )
    ) %>%
    addLegend(
      position = "bottomright",
      pal      = pal_emp,
      values   = ~sqrt(Emp_Tot),
      title    = htmltools::HTML("<b>Total Employment</b><br><span style='font-size:11px;'>Lighter = fewer jobs<br>Darker = more concentrated employment</span>"),
      opacity  = 0.85,
      labFormat = labelFormat(big.mark = ",")
    ) %>%
    addLayersControl(
      baseGroups = c("Topographic (City Labels)", "Street View", "Light Mode"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    setView(lng = -76.7, lat = 39.0, zoom = 8)
})

```

### County-Level Employment Map
```{r}
renderLeaflet({
  input$year_choice
  # Aggregate total employment by county
  zones_df <- zones_reactive()

  emp_by_county <- zones_df %>%
    st_make_valid() %>%
    group_by(County) %>%
    summarise(Total_EMP = sum(Emp_Tot, na.rm = TRUE), .groups = "drop") %>%
    st_union(by_feature = TRUE) %>%
    st_buffer(0) %>%
    st_as_sf()

  # Palette for employment counts
  emp_by_county$EMP_SQRT <- sqrt(emp_by_county$Total_EMP)
  pal_emp_county <- colorNumeric(
    colorRampPalette(c("#fff5f0", "#fcbba1", "#fb6a4a", "#cb181d", "#67000d"))(8),
    domain = emp_by_county$EMP_SQRT,
    na.color = "transparent"
  )

  leaflet(emp_by_county, options = leafletOptions(minZoom = 6, zoomControl = TRUE)) %>%
    addProviderTiles(providers$Esri.WorldTopoMap,   group = "Topographic (City Labels)") %>%
    addProviderTiles(providers$Esri.WorldStreetMap, group = "Street View") %>%
    addProviderTiles(providers$CartoDB.Positron,    group = "Light Mode") %>%
    addPolygons(
      fillColor   = ~pal_emp_county(EMP_SQRT),
      fillOpacity = 0.72,
      color       = "#2b2b2b",
      weight      = 1.1,
      label       = ~sprintf("<b>%s County</b><br><b>Total Employment:</b> %s",
                             County, format(Total_EMP, big.mark = ",")) %>% lapply(htmltools::HTML),
      highlightOptions = highlightOptions(weight = 2, color = "#000000", fillOpacity = 0.9, bringToFront = TRUE)
    ) %>%
    addLegend(
      "bottomright", pal = pal_emp_county, values = ~EMP_SQRT, opacity = 0.85,
      title = htmltools::HTML("<b>Total Employment</b><br><span style='font-size:11px;'>Lighter = lower<br>Darker = higher</span>"),
      labFormat = labelFormat(big.mark = ",")
    ) %>%
    addLayersControl(
      baseGroups = c("Topographic (City Labels)", "Street View", "Light Mode"),
      options = layersControlOptions(collapsed = FALSE)
    ) %>%
    setView(lng = -76.7, lat = 39.0, zoom = 8)
})

```
## Row
### Employment Density by Region
```{r}
renderPlotly({
  input$year_choice
  
  emp_density_region <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(
      Emp_Density = sum(Emp_Tot, na.rm = TRUE) / sum(AREA_KM2, na.rm = TRUE)
    ) %>%
    arrange(desc(Emp_Density))
  
  p_emp_density <- ggplot(emp_density_region, aes(
    x = reorder(REGION, Emp_Density),
    y = Emp_Density,
    fill = REGION,
    text = paste(
      "<b>Region:</b>", REGION,
      "<br><b>Employment Density:</b>", round(Emp_Density, 1), "jobs per km²"
    )
  )) +
    geom_col(width = 0.7, alpha = 0.9) +
    coord_flip() +
    scale_y_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +
    labs(
      title = "Employment Density by Region",
      subtitle = "Average number of jobs per square kilometer",
      x = "Region",
      y = "Employment Density (jobs per km²)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    ) +
    guides(fill = "none")
  
  ggplotly(p_emp_density, tooltip = "text") %>%
    layout(
      title = list(
        text = "<b>Employment Density by Region</b><br><span style='font-size:12px;color:gray'>Jobs per km²</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Employment Density (jobs per km²)", font = list(size = 13)), tickformat = ",f"),
      yaxis = list(title = list(text = "Region", font = list(size = 13)), autorange = "reversed"),
      margin = list(l = 140, r = 50, t = 100, b = 100)
    )
})
```

## Row {.full}
### Employment by Sector and Region
```{r}
renderPlotly({
  input$year_choice
  # Define sector columns and friendly labels
  sector_cols <- c("Emp_Ret", "Emp_Off", "Emp_Ind", "Emp_Edu", "Emp_Hea", "Emp_Foo", "Emp_Oth")
  sector_labels <- c(
    "Emp_Edu" = "Education",
    "Emp_Foo" = "Food & Accommodation",
    "Emp_Hea" = "Healthcare",
    "Emp_Ind" = "Industrial / Manufacturing",
    "Emp_Off" = "Office / Professional Services",
    "Emp_Oth" = "Other Services",
    "Emp_Ret" = "Retail Trade"
  )

  # Summarise employment by sector and region
  emp_long <- st_drop_geometry(zones_reactive()) %>%
    select(REGION, all_of(sector_cols)) %>%
    pivot_longer(cols = all_of(sector_cols), names_to = "Sector", values_to = "Employees") %>%
    mutate(Sector = recode(Sector, !!!sector_labels)) %>%
    group_by(REGION, Sector) %>%
    summarise(Employees = sum(Employees, na.rm = TRUE), .groups = "drop")

  # Create stacked bar chart
  p_emp_sector <- ggplot(emp_long, aes(x = Sector, y = Employees, fill = REGION)) +
    geom_col(position = "stack", width = 0.8) +
    labs(
      title = "Employment by Sector and Region",
      subtitle = "",
      x = "Employment Sector",
      y = "Number of Employees"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      axis.text.x = element_text(angle = 25, vjust = 1, hjust = 1, size = 11, face = "bold")
    )

  # Render interactive plot
  ggplotly(p_emp_sector) %>%
    layout(
      title = list(
        text = "<b>Employment by Sector and Region</b><br><span style='font-size:12px;color:gray'>Total employees by sector across all regions</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Employment Sector", font = list(size = 13))),
      yaxis = list(title = list(text = "Number of Employees", font = list(size = 13)), tickformat = ","),
      margin = list(l = 80, r = 50, t = 90, b = 100)
    )
})

```

## Row
### Total Employment by Region
```{r}
renderPlotly({
  input$year_choice

  # Compute total employment by region
  emp_by_region <- st_drop_geometry(zones_reactive()) %>%
    group_by(REGION) %>%
    summarise(Total_Emp = sum(Emp_Tot, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(Total_Emp))

  p_emp_tot <- ggplot(
    emp_by_region,
    aes(
      x = Total_Emp,
      y = reorder(REGION, Total_Emp),
      fill = REGION,
      text = paste0(
        "<b>Region:</b> ", REGION,
        "<br><b>Total Employment:</b> ", format(Total_Emp, big.mark = ",", scientific = FALSE)
      )
    )
  ) +
    geom_col(alpha = 0.9, width = 0.6, show.legend = TRUE) +
    # ✅ Format axis numbers with commas
    scale_x_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +
    labs(
      title = "Total Employment by Region",
      subtitle = "",
      x = "Total Employment",
      y = NULL
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
      axis.text.y = element_text(size = 12, face = "bold", hjust = 1, margin = margin(r = 10)),
      axis.text.x = element_text(size = 11),
      legend.position = "right",
      legend.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 11),
      plot.margin = margin(t = 20, r = 40, b = 80, l = 180)
    )

  # ✅ Render interactive plot with comma format on axis ticks
  ggplotly(p_emp_tot, tooltip = "text") %>%
    layout(
      title = list(
        text = "<b>Total Employment by Region</b><br><span style='font-size:12px;color:gray'>Regional employment totals</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(
        title = list(text = "Total Employment", font = list(size = 13)),
        tickformat = ",f"  # ✅ comma-formatted values (e.g., 1,000,000)
      ),
      yaxis = list(
        title = list(text = ""),
        automargin = TRUE
      ),
      margin = list(l = 160, r = 40, t = 90, b = 80)
    )
})

```

## Row
### Employment-to-Population Ratio by Region
```{r}
renderPlotly({
  input$year_choice
  # Compute average employment-to-population ratio by region
  emp_ratio_region <- st_drop_geometry(zones_reactive()) %>%
    mutate(Emp_to_Pop_Ratio_Clean = Emp_to_Pop_Ratio / 10) %>%
    group_by(REGION) %>%
    summarise(Avg_Ratio = mean(Emp_to_Pop_Ratio_Clean, na.rm = TRUE)) %>%
    arrange(desc(Avg_Ratio))

  p_emp_ratio <- ggplot(emp_ratio_region, aes(
    x = reorder(REGION, Avg_Ratio),
    y = Avg_Ratio,
    fill = REGION,
    text = paste0(
      "<b>Region:</b> ", REGION,
      "<br><b>Employment-to-Population Ratio:</b> ", round(Avg_Ratio, 1), "%"
    )
  )) +
    geom_col(alpha = 0.9, width = 0.6, show.legend = FALSE) +
    coord_flip() +
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    labs(
      title = "Employment-to-Population Ratio by Region",
      subtitle = "Share of residents engaged in employment (%)",
      x = "Region",
      y = "Employment-to-Population Ratio (%)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
      axis.text.y = element_text(size = 12, face = "bold", hjust = 1, margin = margin(r = 10)),
      axis.text.x = element_text(size = 11)
    )

  # Render interactive plot
  ggplotly(p_emp_ratio, tooltip = "text") %>%
    layout(
      title = list(
        text = "<b>Employment-to-Population Ratio by Region</b><br><span style='font-size:12px;color:gray'>Average share of working residents per region</span>",
        xanchor = "center", x = 0.5
      ),
      xaxis = list(title = list(text = "Employment-to-Population Ratio (%)", font = list(size = 13))),
      yaxis = list(title = list(text = "Region", font = list(size = 13)), autorange = "reversed"),
      margin = list(l = 140, r = 40, t = 100, b = 120)
    )
})

```

## Row
### Employment by Sector (Total)
```{r}
renderPlotly({
  input$year_choice
  
  # Summarise employment totals across all sectors
  sector_cols <- c("Emp_Ret", "Emp_Off", "Emp_Ind", "Emp_Edu", "Emp_Hea", "Emp_Foo", "Emp_Oth")

  emp_sector_tot <- st_drop_geometry(zones_reactive()) %>%
    summarise(across(all_of(sector_cols), ~ sum(.x, na.rm = TRUE))) %>%
    pivot_longer(everything(), names_to = "Sector", values_to = "Total_Employees") %>%
    mutate(Sector = gsub("Emp_", "", Sector))

  # Replace abbreviated sector names with human-readable labels
  emp_sector_tot <- emp_sector_tot %>%
    mutate(Sector = recode(Sector,
      "Edu" = "Education",
      "Foo" = "Food & Accommodation",
      "Hea" = "Healthcare",
      "Ind" = "Industrial / Manufacturing",
      "Off" = "Office / Professional Services",
      "Oth" = "Other Services",
      "Ret" = "Retail Trade",
      "to_Pop_Ratio" = "Employment-to-Population Ratio"
    ))

  # Build chart
  p_emp_sector_tot <- ggplot(
    emp_sector_tot,
    aes(
      x = reorder(Sector, Total_Employees),
      y = Total_Employees,
      fill = Sector,
      text = paste0(
        "<b>Sector:</b> ", Sector,
        "<br><b>Total Employment:</b> ", format(Total_Employees, big.mark = ",", scientific = FALSE)
      )
    )
  ) +
    geom_col(show.legend = FALSE, alpha = 0.9) +
    coord_flip() +
    # ✅ Add comma formatting to numeric axis
    scale_y_continuous(labels = function(x) format(round(x, 0), big.mark = ",", scientific = FALSE)) +
    labs(
      title = "Employment by Sector",
      subtitle = "",
      x = "Employment Sector",
      y = "Total Employees"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30")
    )

  # Render interactive plot if data exists
  if (nrow(emp_sector_tot) > 0) {
    ggplotly(p_emp_sector_tot, tooltip = "text") %>%
      layout(
        title = list(
          text = "<b>Employment by Sector</b><br><span style='font-size:12px;color:gray'>Total employees across all sectors</span>",
          xanchor = "center", x = 0.5
        ),
        # ✅ comma-formatted x-axis ticks
        xaxis = list(
          title = list(text = "Total Employees", font = list(size = 13)),
          tickformat = ",f"
        ),
        yaxis = list(title = list(text = "Employment Sector", font = list(size = 13))),
        margin = list(l = 140, r = 50, t = 100, b = 100)
      )
  }
})

```
